---
title: "BigMart Sales Analysis"
author: "Ananya Jha"
output: html_document
---

```{r}
library(tidyverse)
library(here)
library(ggplot2)
library(ggthemes)
```

# 1. Introduction

The goal of this project is to uncover business insights by analyzing retail sales from BigMart outlets and identifying the key factors influencing product sales.

The analysis focuses on:

* Sales distribution patterns
* Impact of pricing and visibility
* Outlet performance comparison
* Product category contribution

---

# 2. Data Overview

```{r}
data <- read_csv(here("data", "cleaned", "bigmart_cleaned.csv"))

glimpse(data)
```

The dataset contains over 8,500 observations with multiple features related to product characteristics, outlet details, and sales performance.

---

# 3. Data Cleaning Summary

The dataset required several preprocessing steps:

* Missing values in `item_weight` were imputed using group mean
* Missing `outlet_size` replaced with "Unknown"
* Zero values in `item_visibility` were imputed using group median
* Standardized `item_fat_content` categories
* Converted multiple "character" data types into "factors"
* Created new features:

  * `outlet_age`
  * `price_category`

---

### Note: Exploratory Analysis Approach

This report presents a selection of key exploratory analyses and visualizations.
The complete analysis, including detailed EDA and intermediate outputs, is available in the scripts/ directory.

The focus of this report is to highlight the most relevant insights and business-critical patterns derived from the data.


# 4. Exploratory Data Analysis (EDA)

### 4.1 Univariate Analysis

```r
summary(data$item_outlet_sales)

data %>%
  ggplot(aes(x = item_outlet_sales)) +
  geom_histogram(fill = "#8A226AFF", bins = 50) + 
  labs(
    title = "Sales Distribution",
    subtitle = "Majority of Products cluster in lower sales range, with a few high-performing outliers",
    x = "Item Outlet Sales",
    y = "Number of Products",
  ) + 
  theme_fivethirtyeight() +
  theme(
    text = element_text(family = "sans"),
    plot.title = element_text(face = "bold", size = 18),
    axis.title = element_text(size = 14),
    axis.text = element_text()
  )
```

The sales variable shows a wide distribution with a right-skew, indicating that a small number of products generate significantly higher revenue compared to the majority.

---

### 4.2 Bivariate Analysis

#### Sales by Outlet Type

```r
sales_by_outlet <- data %>%
  group_by(outlet_type) %>%
  summarise(avg_sales = mean(item_outlet_sales, na.rm = TRUE),
            total_sales = sum(item_outlet_sales)) %>%
  ungroup()

```

Different outlet types show significant variation in total sales, suggesting that store format plays a key role in performance.

---

#### Sales by Location

```r
sales_by_location <- data %>%
  group_by(outlet_location_type) %>%
  summarise(avg_sales = mean(item_outlet_sales, na.rm = TRUE),
            total_sales = sum(item_outlet_sales)) %>%
  ungroup()
```

Sales patterns differ across location tiers, indicating geographical influence on revenue generation.

---

### 4.3 Multivariate Analysis

#### Outlet Age vs Sales

```r
cor(data$outlet_age, data$item_outlet_sales)
```

The correlation between outlet age and sales is weak, suggesting that older outlets do not necessarily perform better than newer ones.

---

# 5. Visualizations

### 1. Price vs Sales

```{r}
data %>%
  ggplot(aes(x = item_mrp, y = item_outlet_sales)) +
  geom_point(alpha = 0.5, color = "#F9C932FF") + 
  geom_smooth(method = "lm", color = "#8A226AFF") + 
  labs(
    title = "Sales vs MRP",
    subtitle = "Higher item price points indicate positive correlation with total outlet revenue",
    x = "Item MRP",
    y = "Outlet Sales"
  ) +
  theme_fivethirtyeight() +
  theme(
    text = element_text(family = "sans"),
    plot.title = element_text(face = "bold", size = 18),
    axis.title = element_text(size = 14),
    axis.text = element_text()
  )
```

Higher-priced items tend to generate higher sales, indicating a positive correlation between price and revenue.

---

### 2. Visibility vs Sales

```{r}
data %>%
  ggplot(aes(x = item_visibility, y = item_outlet_sales)) +
  geom_point(alpha = 0.3, color = "#F98C0AFF") + 
  geom_smooth(method = "lm", color = "#BA3655FF") +
  labs(
    title = "Visibility vs Sales",
    subtitle = "High visibility items do not consistently outperform low visibility ones",
    x = "Visibility",
    y = "Outlet Sales"
  ) + 
  theme_fivethirtyeight() +
  theme(
    text = element_text(family = "sans"),
    plot.title = element_text(face = "bold", size = 18),
    axis.title = element_text(size = 14),
    axis.text = element_text()
  )
```

Item visibility does not show a strong or consistent impact on sales performance.

---

### 3. Outlet Age vs Sales

```{r}
data %>%
  ggplot(aes(x = factor(outlet_age), y = item_outlet_sales)) +
  geom_boxplot(fill = "#C44E52FF", alpha = 0.9) +
  labs(
    title = "Outlet Age vs Outlet Sales",
    subtitle = "Sales variability remains similar across outlet ages",
    x = "Outlet Age",
    y = "Outlet Sales"
  ) +
  theme_fivethirtyeight() + 
  theme(
    text = element_text(family = "sans"),
    plot.title = element_text(face = "bold", size = 18),
    axis.title = element_text(size = 14),
    axis.text = element_text()
  )
```

Sales distribution appears relatively consistent across outlet ages, suggesting limited impact of outlet age on performance.

---

### 4. Sales by Outlet Type

```{r}
sales_by_outlet <- read_csv(here("data", "processed", "sales_by_outlet.csv"))

sales_by_outlet %>%
  ggplot(aes(x = reorder(outlet_type, total_sales), y = total_sales)) + 
  geom_col(fill = "#F98C0AFF") + 
  scale_y_continuous(labels = scales::comma) + 
  labs(
    title = "Total Sales by Outlet Type",
    subtitle = "Supermarket Type 1 generates the most revenue, significantly outperforming other outlets",
    x = "Outlet Type",
    y = "Outlet Sales"
  ) + 
  theme_fivethirtyeight() + 
  theme(
    text = element_text(family = "sans"),
    plot.title = element_text(size = 18, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text()
  )
```

Supermarket Type 1 generates the highest total sales, significantly outperforming other outlet types.

---

### 5. Sales by Location Tier

```{r}
sales_by_location <- read_csv(here("data", "processed", "sales_by_location.csv"))

sales_by_location %>%
  ggplot(aes(x = outlet_location_type, y = total_sales)) + 
  geom_col(fill = "#57106EFF") + 
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Sales by Location Type",
    subtitle = "Tier 3 locations appears to generate highest sales, suggesting a strong market presence in suburban or rural zones",
    x = "Location Type",
    y = "Total Sales"
  ) + 
  theme_fivethirtyeight() + 
  theme(
    text = element_text(family = "sans"),
    plot.title = element_text(size = 18, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(),
    plot.margin = margin(10, 30, 10, 10)
  )
```

Tier 3 locations generate the highest sales, indicating strong performance in non-urban markets.

---

### 6. Sales by Category

```{r}
sales_by_category <- read_csv(here("data", "processed", "sales_by_category.csv"))

sales_by_category %>%
  ggplot(aes(x = reorder(item_type, total_sales), y = total_sales)) + 
  geom_col(fill = "#BA3655FF") + 
  scale_y_continuous(labels = scales::comma) +
  coord_flip() +
  labs(
    title = "Sales by Category",
    subtitle = "Fruits & Vegetables and Snack Foods dominate total sales contribution",
    x = "Item Type",
    y = "Total Sales"
  ) + 
  theme_fivethirtyeight() + 
  theme(
    text = element_text(family = "sans"),
    plot.title = element_text(size = 18, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text()
  )
```

Fruits & Vegetables and Snack Foods dominate total sales contribution.

---

### 7. Sales by Fat Content

```{r}
sales_by_fat <- read_csv(here("data", "processed", "sales_by_fat.csv"))

sales_by_fat %>%
  ggplot(aes(x = item_fat_content, y = total_sales)) + 
  geom_col(fill = "#8A226AFF") + 
  scale_y_continuous(labels = scales::comma) + 
  coord_flip() + 
  labs(
    title = "Sales by Fat Content",
    subtitle = "Low-fat products account for a larger portion of total sales compared to regular-fat alternatives",
    x = "Fat",
    y = "Sales"
  ) + 
  theme_fivethirtyeight() + 
  theme(
    text = element_text(family = "sans"),
    plot.title = element_text(size = 18, face = "bold"),
    axis.title = element_text(size =14),
    axis.text = element_text()
  )
```

Low-fat products contribute more to total sales than regular-fat products.

---

# 6. Key Insights

* Sales are heavily skewed toward a small number of high-performing products
* Higher MRP items tend to generate more revenue
* Visibility has limited impact on sales performance
* Supermarket Type 1 dominates total sales
* Tier 3 locations outperform urban areas
* Certain product categories contribute disproportionately to revenue

---

# 7. Business Recommendations

* Focus inventory on high-performing product categories
* Expand operations in Tier 3 locations
* Optimize pricing strategies for premium products
* Improve performance of underperforming outlet types

---

# 8. Conclusion

This analysis highlights key drivers of retail performance and provides actionable insights for improving sales strategy and operational efficiency across BigMart outlets.

````